Removing nonlinear batch effects using adversarial deep neural networks

### 2019-04-30 thesis.tex
Figure references, instructions for pdf to docx

\documentclass[notitlepage]{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage[square,numbers]{natbib}
\usepackage[nomarkers,nolists]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\linespread{1.5}
\bibliographystyle{unsrtnat}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
	\caption{\textbf{Batch adjustment justification and steps}
	(a) When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
	\label{fig:workflow}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network}
	\caption{\textbf{Network architecture} of Confounded.
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{equation}
	\label{dual_loss}
	L_dual = L_A - \lambda{}(L_D)
\end{equation}

The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_dual$.

@@

In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
In each iteration, both $L_D$ and $L_dual$ were optimized.
When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca}
	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne}
	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse}
	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
	MSE for unadjusted data will always be 0 because the input data is identical to the output data.
	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mse}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd}
	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mmd}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
	See also Figure \ref{fig:batch}.}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
	See also Figure \ref{fig:true_class}.}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/mnist}
	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,
	(b) with artificial noise added,
	(c) adjusted for noise by ComBat, and
	(d) adjusted for noise by Confounded.
	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}

>>>>

\documentclass[notitlepage]{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage[square,numbers]{natbib}
% \usepackage{natbib}
\usepackage[nomarkers,nolists]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\linespread{1.5}
\bibliographystyle{unsrtnat}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption{\textbf{Batch adjustment justification and steps}
	(a) When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
	\label{fig:workflow}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption{\textbf{Network architecture} of Confounded.
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{equation}
	\label{dual_loss}
	L_{dual} = L_A - \lambda{}(L_D)
\end{equation}

The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the data.
$L_A$ was not optimized directly; instead the autoencoder was trained by optimizing $L_{dual}$.

@@

In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
In each iteration, both $L_D$ and $L_{dual}$ were optimized.
When optimizing $L_D$ (i.e. training the discriminator), the autoencoder's weights were frozen, and vice versa.
For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/training_loss.png}
	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
	MSE for unadjusted data will always be 0 because the input data is identical to the output data.
	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mse}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mmd}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
	See also Figure \ref{fig:batch}.}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
	See also Figure \ref{fig:true_class}.}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/mnist.png}
	\caption{\textit{TODO: this is old output data and needs to be updated.} \textbf{MNIST handwritten digits} (a) before any adjustment,
	(b) with artificial noise added,
	(c) adjusted for noise by ComBat, and
	(d) adjusted for noise by Confounded.
	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}
